Human capital mediates natural selection in contemporary humans

### 2022-05-15 why-natural-selection.Rmd
Further graphical improvements:   - use color and fill for p values   - fewer grid lines   - many tweaks   - remove cruft

  if (conf_int) {
    confint_scale <- scale_linetype_manual(
                       name   = "", 
                       values = c("95% c.i. uncorrected" = 1),
                       guide  = guide_legend(order = 3)
                     )
    confint_geom_segment <- confint_geom_segment()
  } else {
    confint_scale <- NULL
    confint_geom_segment <- NULL
  }

@@

  n_regs <- as.double(n_regs)
  shape_values <- c(16, 15, 17, 18, 20)[1:n_cats]
  if (fill_direction == -1) shape_values <- rev(shape_values)
  dfr %>% mutate(
    !! score_col := pretty_names(!! score_col),
    !! score_col := fct_reorder(!! score_col, estimate, order_abs(order_idx))
  ) %>% 
    rename(p = p.value) %>% 
    ggplot(aes(estimate, {{score_col}}, yend = {{score_col}}, 
               shape = {{fill_col}}, ...)) +
    my_vline + 
    confint_geom_segment +
    geom_point(size = 1.8, alpha = 0.9, aes(color = {{fill_col}})) +
    geom_point(size = 0.7, color = "white", aes(alpha = p < 0.05/{{n_regs}})) +
    my_fill_scale(aesthetics = "color", n = n_cats, direction = fill_direction, 
                    guide = my_shape_fill_guide) +
    scale_shape_manual(values = shape_values, guide = my_shape_fill_guide) +
    my_labs() +
    my_theme(legend.justification = "top") +
    confint_scale +
    scale_alpha_discrete(range = c(1, 0), 
                           guide = guide_legend(
                             order = 2,
                             override.aes = list(
                               shape = c(1, 16),
                               color = "black",
                               alpha = 1,
                               size = 1.8
                             )
                         )) 
}

my_theme <- function (...) {
  theme(
          axis.text.y = element_text(size = 7), 
          ...
        )
}

confint_geom_segment <- function () {
  geom_segment(aes(x = conf.low, xend = conf.high, linetype = "95% c.i. uncorrected"), 
                    color = "grey55", alpha = 0.2, size = 1.1, 
                    lineend = "round")
}

@@

dfr %>% 
      mutate(
        PGS      = ifelse(abs(effect_size) > 0.027, PGS, ""),
        PGS      = sub("(^\\w*?)_.*", "\\1", PGS)
      ) %>% 
      filter(
        Controls == "None"
      ) %>% 
      ggplot(aes(effect_size, cor)) + 
        geom_point(color = "steelblue4") +
        geom_text(aes(label = PGS), size = 3, colour = "black", 
                    check_overlap = TRUE, nudge_x = 0.002) +
        facet_wrap( ~ Correlation, scales = "free") + 
        my_vline + 
        my_hline +
        labs(x = "Selection effect", y = "Correlation") +
        coord_cartesian(clip = "off") +
        theme(
          legend.position = "none",
          panel.spacing = unit(20, "pt")
        )

@@

res_mediation %>% 
      mutate(
        term = pretty_names(term),
        term = fct_reorder(term, prop_ind, order_abs(1)),
        `Indirect effect p < 0.05/23` = sig 
      ) %>% 
      ggplot(aes(y = term)) + 
        geom_col(aes(x = prop_ind, fill = `Indirect effect p < 0.05/23`), 
                   width = .5) + 
        geom_errorbar(aes(
                          xmin = prop_ind_conf_low, 
                          xmax = prop_ind_conf_high,
                          linetype = "Bootstrap 95% c.i. uncorrected",
                        ),
                        width = 0.2,
                        color = "grey50"
                      ) +
        scale_fill_manual(values = c("FALSE" = "grey65", "TRUE" = "blue3")) +
        my_vline +
        scale_x_continuous(labels = scales::percent, n.breaks = 7) +
        coord_cartesian(xlim = c(-0.5, 1)) +
        labs(x = "Proportion mediated", y = "") +
        scale_linetype_manual(name = "", values = c("Bootstrap 95% c.i. uncorrected" = 1))

@@

Because selection effects are concentrated in lower-income groups, they may also
increase inequality with respect to polygenic scores. For example, Figure
\@ref(fig:plot-mean-EA3-by-income) plots mean polygenic scores for educational
attainment (EA3) among children from households of different income groups. The
red bars show the actual means, i.e. parents' mean polygenic score weighted by
number of children. The grey bars show the hypothetical means if all households
had equal numbers of children. Natural selection against genes associated with
educational attainment is stronger at the bottom of the income distribution, and
this increases the differences between groups. Overall, natural selection
increases the correlation of polygenic scores with income for `r n_increases`
out of 33 polygenic scores, with a median percentage increase of 
`r median_pct_change`% in the respondents' generation (Appendix Table
\@ref(tab:tbl-inequality-effects)).  If inequalities in polygenic scores are
important for understanding social structure and mobility [@belsky2018genetic;
@Rimfeld_2018; @harden2021genetic], then these increases are substantive.
Similarly, since many polygenic scores are predictive of disease risk, they
could potentially increase health inequalities. In general, the evolutionary
history of anatomically modern humans is related to disease risk
[@benton2021influence]; understanding the role of contemporary natural selection
may help researchers to map the genetic architecture of health disparities.

```{r plot-mean-EA3-by-income, fig.align = "center", fig.cap = "Mean polygenic score for educational attainment (EA3) of children by household income group. Blue is actual. Grey is hypothetical in the absence of selection effects.", fig.width = 4}
income_EA3 <- famhist %>% 
      filter(! is.na(n_children), ! is.na(income_cat)) %>% 
      mutate(
        `Household income` = factor(income_cat, 
                          labels = c("< £18K", "£18-30K", "£31-51K", "£52-100K", 
                                      "> £100K"))
      ) %>% 
      group_by(`Household income`) %>% 
      summarize(
        `Without selection` = mean(EA3_excl_23andMe_UK, na.rm = TRUE),
        Actual              = weighted.mean(EA3_excl_23andMe_UK, n_children, 
                                              na.rm = TRUE)
      ) %>% 
      tidyr::pivot_longer(-`Household income`) %>% 
      rename(
        `Children's mean EA3` = value
      )

income_EA3 %>% 
              ggplot(aes(`Household income`, `Children's mean EA3`, fill = name)) +
                geom_col(position = "dodge") +
                scale_fill_manual(values = c("darkred", "grey65")) +
                theme(
                  legend.background = element_rect(fill = "white", colour = NA),
                  legend.title = element_blank(), 
                  legend.position = c(.25, .85)
                )
```

@@

```{r plot-age-birth-parents-cross, fig.cap = "Selection effects (parents' RLRS) by age at first live birth terciles.  Weights multiplied by 1/number of siblings.", fig.subcap = c("Mothers", "Fathers"), fig.ncol = 1}

@@

```{r plot-age-birth-parents, fig.cap = "Selection effects (parents' generation) among eldest siblings, controlling for parents' age at birth. Weights multiplied by 1/number of siblings."}

>>>>

  if (conf_int) {
    confint_scale <- scale_linetype_manual(
                       name   = "", 
                       values = c("95% c.i. uncorrected" = 1),
                       guide  = guide_legend(order = 3)
                     )
    confint_geom_segment <- geom_segment(
                              aes(
                                x        = conf.low, 
                                xend     = conf.high,
                                linetype = "95% c.i. uncorrected"
                              ), 
                              color   = "grey45", 
                              alpha   = 0.2, 
                              size    = 1.1, 
                              lineend = "round"
                            )
  } else {
    confint_scale <- NULL
    confint_geom_segment <- NULL
  }

@@

  n_regs <- as.double(n_regs)
  shape_values <- paste(c("circle", "square", "triangle", "diamond", 
                            "triangle down"), "filled")
  shape_values <- shape_values[1:n_cats]
  if (fill_direction == -1) shape_values <- rev(shape_values)
  dfr %>% mutate(
    !! score_col := pretty_names(!! score_col),
    !! score_col := fct_reorder(!! score_col, estimate, order_abs(order_idx))
  ) %>% 
    rename(p = p.value) %>% 
    ggplot(aes(estimate, {{score_col}}, yend = {{score_col}}, 
               shape = {{fill_col}}, ...)) +
    my_vline + 
    confint_geom_segment +
    geom_point(size = 1.8, alpha = 0.8, aes(color = {{fill_col}}, 
                 fill = stage(p < 0.05/{{n_regs}}, after_scale = ifelse(
                 fill == "white", fill, color)))) +
    my_fill_scale(aesthetics = "color", n = n_cats, direction = fill_direction, 
                    guide = my_shape_fill_guide) +
    scale_fill_manual(values = c("TRUE" = "black", "FALSE" = "white"), 
                        guide = guide_legend(
                                  title = sprintf("p < 0.05/%s", n_regs),
                                  override.aes = list(shape = "circle filled"),
                                  order = 2
                                )) + 
    scale_shape_manual(values = shape_values, guide = my_shape_fill_guide) +
    my_labs() +
    my_theme(
      legend.justification = "top", 
      panel.grid.major.y   = element_blank()
    ) +
    confint_scale
}

my_theme <- function (...) {
  theme(
          axis.text.y        = element_text(size = 7),
          panel.grid.major   = element_line(size = rel(0.5)),
          ...
        )
}

@@

dfr %>% 
      mutate(
        PGS      = ifelse(abs(effect_size) > 0.027, PGS, ""),
        PGS      = sub("(^\\w*?)_.*", "\\1", PGS)
      ) %>% 
      filter(
        Controls == "None"
      ) %>% 
      ggplot(aes(effect_size, cor)) + 
        geom_point(color = "steelblue4") +
        geom_text(aes(label = PGS), size = 3, colour = "black", 
                    check_overlap = TRUE, nudge_x = 0.008) +
        facet_wrap( ~ Correlation, scales = "free") + 
        my_vline + 
        my_hline +
        labs(x = "Selection effect", y = "Correlation") +
        coord_cartesian(clip = "off") +
        theme(
          legend.position = "none",
          panel.spacing = unit(20, "pt")
        )

@@

res_mediation %>% 
      mutate(
        term = pretty_names(term),
        term = fct_reorder(term, prop_ind, order_abs(1)),
        `Indirect effect p < 0.05/23` = sig 
      ) %>% 
      ggplot(aes(y = term)) + 
        geom_col(aes(x = prop_ind, fill = `Indirect effect p < 0.05/23`), 
                   width = .5) + 
        geom_errorbar(aes(
                          xmin = prop_ind_conf_low, 
                          xmax = prop_ind_conf_high,
                          linetype = "Bootstrap 95% c.i. uncorrected",
                        ),
                        width = 0.2,
                        color = "grey50"
                      ) +
        scale_fill_manual(values = c("FALSE" = "grey65", "TRUE" = "darkgreen")) +
        my_vline +
        scale_x_continuous(labels = scales::percent, n.breaks = 7) +
        coord_cartesian(xlim = c(-0.5, 1)) +
        labs(x = "Proportion mediated", y = "") +
        scale_linetype_manual(name = "", values = c("Bootstrap 95% c.i. uncorrected" = 1)) +
        my_theme(
          panel.grid.major.y = element_blank(),
          panel.grid.minor.x = element_blank()
        )

@@

Because selection effects are concentrated in lower-income groups, they may also
increase inequality with respect to polygenic scores. For example, Figure
\@ref(fig:plot-mean-EA3-by-income) plots mean polygenic scores for educational
attainment (EA3) among children from households of different income groups. The
green bars show the actual means, i.e. parents' mean polygenic score weighted by
number of children. The grey bars show the hypothetical means if all households
had equal numbers of children. Natural selection against genes associated with
educational attainment is stronger at the bottom of the income distribution, and
this increases the differences between groups. Overall, natural selection
increases the correlation of polygenic scores with income for `r n_increases`
out of 33 polygenic scores, with a median percentage increase of 
`r median_pct_change`% in the respondents' generation (Appendix Table
\@ref(tab:tbl-inequality-effects)).  If inequalities in polygenic scores are
important for understanding social structure and mobility [@belsky2018genetic;
@Rimfeld_2018; @harden2021genetic], then these increases are substantive.
Similarly, since many polygenic scores are predictive of disease risk, they
could potentially increase health inequalities. In general, the evolutionary
history of anatomically modern humans is related to disease risk
[@benton2021influence]; understanding the role of contemporary natural selection
may help researchers to map the genetic architecture of health disparities.

```{r plot-mean-EA3-by-income, fig.align = "center", fig.cap = "Mean polygenic score for educational attainment (EA3) of children by household income group. Green is actual. Grey is hypothetical in the absence of selection effects.", fig.width = 4}
income_EA3 <- famhist %>% 
      filter(! is.na(n_children), ! is.na(income_cat)) %>% 
      mutate(
        `Household income` = factor(income_cat, 
                          labels = c("< £18K", "£18-30K", "£31-51K", "£52-100K", 
                                      "> £100K"))
      ) %>% 
      group_by(`Household income`) %>% 
      summarize(
        `Without selection` = mean(EA3_excl_23andMe_UK, na.rm = TRUE),
        Actual              = weighted.mean(EA3_excl_23andMe_UK, n_children, 
                                              na.rm = TRUE)
      ) %>% 
      tidyr::pivot_longer(-`Household income`) %>% 
      rename(
        `Children's mean EA3` = value
      )

income_EA3 %>% 
              ggplot(aes(`Household income`, `Children's mean EA3`, fill = name)) +
                geom_col(position = "dodge") +
                scale_fill_manual(values = c("darkgreen", "grey65")) +
                theme(
                  legend.background = element_rect(
                                        fill = "white", 
                                        colour = "black",
                                        size = rel(0.5)
                                      ),
                  legend.margin = margin(t = 0, b = 5.5, l = 5.5, 
                                           r = 5.5),
                  legend.title = element_blank(), 
                  legend.position = c(.25, .85)
                )
```

@@

```{r plot-age-birth-parents-cross, fig.cap = "Selection effects (parents' RLRS) among eldest siblings, by parents' age at first live birth terciles.  Weights multiplied by 1/number of siblings.", fig.subcap = c("Mothers", "Fathers"), fig.ncol = 1}

@@

```{r plot-age-birth-parents, fig.cap = "Selection effects (parents' RLRS) among eldest siblings, controlling for parents' age at birth. Weights multiplied by 1/number of siblings."}
